Staged Abstract Interpreters

### 2019-09-11 paper/20_prelim.tex
polish
lambdas

\paragraph{The Interpreter} We can now construct the semantics-agnostic interpreter
in monadic form, as shown in Figure \ref{fig:shared_int}. The basic idea
of generic interpretation is to traverse the abstract syntax tree
while maintaining the effects (such as environment and state updates).
Note that the interpreter is written in open-recursive style: it cannot
refer to itself directly. Instead, @eval@ takes an additional parameter @ev@ of
type @Expr => Ans@ that refers to itself. Consequently, the method @close@ that
lifts lambda terms to closures and method @ap_clo@ that applies functions also
takes an extra @ev@, because evaluations may happen inside them.
To close the open recursion, we use a fixed-point operator @fix@.
For concrete-interpretation instantiation, @fix@ works like the Y combinator;
for abstract-interpretation instantiation, it will instrument the interpreter
by memoizing @ev@'s inputs and outputs, which ensures the termination of
abstract interpretation.
Finally, there is a top-level wrapper @run@. The return type @Result@
depends on the kind of monads being used and is therefore also left abstract.
\begin{lstlisting}
  def fix(ev: (Expr => Ans) => (Expr => Ans)): Expr => Ans
  type Result; def run(e: Expr): Result
\end{lstlisting}

@@

\paragraph{Concrete Components}
The two types we need to concretize are addresses @Addr@ and values @Value@. The
types @Env@ and @Store@ are derived automatically. To ensure the freshness of
address allocations, we use type @Int@ and always return a number that is greater
than the size of the current store.
A value can be either a tagged number \texttt{IntV}, or a closure
\texttt{CloV} that contains a lambda term and an environment. The final
result of the interpreter is a pair of values and stores.
To distinguish from the monadic values of type @Ans@ produced by the
interpreter, later we use term \textit{grounded values} to denote such final values.
The two elements in the type @Result@ are annotated by the binding-time @R@,
because they can be next-stage objects.  We also define a standard fixed-point
combinator to close the open-recursive function @ev@.
\begin{lstlisting}
  trait ConcreteComponents extends Semantics {
    type Addr = Int;  sealed trait Value
    case class IntV(i: Int) extends Value
    case class CloV(λ: Lam, e: Env) extends Value
    type Result = (R[Value], R[Store])
    def fix(ev: (Expr => Ans) => (Expr => Ans)): Expr => Ans = e => ev(fix(ev))(e)
  }
\end{lstlisting}

@@

\paragraph{Primitive Operations}
Other primitive operations over the value domains can be implemented
straightforwardly. We elide most of them but describe a bit on how we handle
functions and applications. The reason is that lambda terms are data but are also
part of the control flow. The way we treat them is simple now, but
will become more involved when we stage the abstract interpreter.
The method @close@ denotes a lambda term to our representation of functions.
At the moment, @close@ takes a lambda term and an environment, and
produces the defunctionalized representation of closures.
The method @ap_clo@ performs function applications, which takes a function
value @fun@ and an argument @arg@. It first extracts the lambda term and environment
enclosed in @fun@, and allocates an addresses for the argument and updates the
environment and store accordingly.  Finally, body expression of the lambda term
is evaluated under the new environment and store.
\begin{lstlisting}
  def close(ev: Expr => Ans)(λ: Lam, ρ: Env): Value = CloV(λ, ρ)
  def ap_clo(ev: Expr => Ans)(fun: Value, arg: Value): Ans = fun match {
    case CloV(Lam(x, e), ρ: Env) => for {
      α <- alloc(x)
      _ <- set_store(α → arg)
      rt <- local_env(ev(e))(ρ + (x → α))
    } yield rt
  }
\end{lstlisting}

>>>>

\paragraph{The Interpreter} We can now construct the semantics-agnostic interpreter
in monadic form, as shown in Figure \ref{fig:shared_int}. The basic idea
of generic interpretation is to traverse the AST
while maintaining the effects (such as environment and state updates).
Note that the interpreter is written in open-recursive style: it cannot
refer to itself directly. Instead, @eval@ takes an additional parameter @ev@ of
type @Expr => Ans@ that refers to itself. Consequently, the method @close@ that
lifts $\lambda$-terms to closures and method @ap_clo@ that applies functions also
takes an extra @ev@, because evaluations may happen inside them.
To close the open recursion, we use a fixed-point operator @fix@.
For concrete-interpretation instantiation, @fix@ works like the Y combinator;
for abstract-interpretation instantiation, it will instrument the interpreter
by memoizing @ev@'s inputs and outputs, which ensures the termination of
abstract interpretation.
Finally, there is a top-level wrapper @run@. The return type @Result@
depends on the kind of monads being used and is therefore also left abstract.
\begin{lstlisting}
  def fix(ev: (Expr => Ans) => (Expr => Ans)): Expr => Ans
  type Result; def run(e: Expr): Result
\end{lstlisting}

@@

\paragraph{Concrete Components}
The two types we need to concretize are addresses @Addr@ and values @Value@. The
types @Env@ and @Store@ are derived automatically. To ensure the freshness of
address allocations, we use type @Int@ and always return a number that is greater
than the size of the current store.
A value can be either a tagged number \texttt{IntV}, or a closure
\texttt{CloV} that contains a $\lambda$-term and an environment. The final
result of the interpreter is a pair of values and stores.
To distinguish from the monadic values of type @Ans@ produced by the
interpreter, later we use term \textit{grounded values} to denote such final values.
The two elements in the type @Result@ are annotated by the binding-time @R@,
because they can be next-stage objects.  We also define a standard fixed-point
combinator to close the open-recursive function @ev@.
\begin{lstlisting}
  trait ConcreteComponents extends Semantics {
    type Addr = Int;  sealed trait Value
    case class IntV(i: Int) extends Value
    case class CloV(λ: Lam, e: Env) extends Value
    type Result = (R[Value], R[Store])
    def fix(ev: (Expr => Ans) => (Expr => Ans)): Expr => Ans = e => ev(fix(ev))(e)
  }
\end{lstlisting}

@@

\paragraph{Primitive Operations}
Other primitive operations over the value domains can be implemented
straightforwardly. We elide most of them but describe a bit on how we handle
functions and applications. The reason is that $\lambda$-terms are data but are also
part of the control flow. The way we treat them is simple now, but
will become more involved when we stage the abstract interpreter.
The method @close@ denotes a $\lambda$-term to our representation of functions.
At the moment, @close@ takes a $\lambda$-term and an environment, and
produces the defunctionalized representation of closures.
The method @ap_clo@ performs function applications, which takes a function
value @fun@ and an argument @arg@. It first extracts the syntactic
$\lambda$-term and environment enclosed in @fun@, and allocates an addresses
for the argument and updates the environment and store accordingly.  Finally,
body expression of the $\lambda$-term is evaluated under the new environment
and store.
\begin{lstlisting}
  def close(ev: Expr => Ans)(λ: Lam, ρ: Env): Value = CloV(λ, ρ)
  def ap_clo(ev: Expr => Ans)(fun: Value, arg: Value): Ans = fun match {
    case CloV(Lam(x, e), ρ: Env) => for {
      α <- alloc(x)
      _ <- set_store(α → arg)
      rt <- local_env(ev(e))(ρ + (x → α))
    } yield rt
  }
\end{lstlisting}

### 2019-09-11 paper/30_stageconc.tex
space
lambdas

\paragraph{Primitive Operations} Most of the primitive operations can be easily
translated to their staged versions -- we just need to change the types.  As we
mentioned before, we will illustrate in detail how functions and applications
are handled.  One problem here is what should we do for $\lambda$ terms when
staging them? We cannot directly create a next-stage defunctionalized closure
for it, because that means we still need to interpret over these closures at the
next stage and the interpretation overhead has not been eliminated. 
The solution is to compile the $\lambda$ term with its environment by calling
the evaluator at the current stage. The result is generating a next-stage
Scala function. The evaluator is passed as @ev@ to the method @close@.
The following code implements the idea:
\begin{lstlisting}
  type ValSt = (Value, Store)
  def emit_compiled_clo(f: (Rep[Value], Rep[Store]) => Rep[ValSt], λ: Lam, ρ: Exp[Env]): Rep[Value]
  def close(ev: Expr => Ans)(λ: Lam, ρ: Rep[Env]): Rep[Value] = {
    val Lam(x, e) = λ
    val f: (Rep[Value], Rep[Store]) => Rep[ValSt] = { case (v: Rep[Value], σ: Rep[Store]) =>
        val α = alloc(σ, x)
        ev(e)(ρ + (unit(x) → α))(σ + (α → v))
    }; emit_compiled_clo(f, λ, ρ)
  }
\end{lstlisting}

@@

Now, compared with the unstaged interpreter, we observe a key difference of where
we use the evaluation function @ev@.  In the unstaged interpreter, the
invocation of @ev@ happens in @ap_clo@, i.e., at the application time; while in
the staged interpreter, we eagerly apply @ev@ when denoting the lambda terms to
our next-stage value domains, i.e., at the value-representation time.

>>>>

\paragraph{Primitive Operations} Most of the primitive operations can be easily
translated to their staged versions -- we just need to change the types.  As we
mentioned before, we will illustrate in detail how functions and applications
are handled.  One problem here is what should we do for $\lambda$-terms when
staging them? We cannot directly create a next-stage defunctionalized closure
for it, because that means we still need to interpret over these closures at the
next stage and the interpretation overhead has not been eliminated. 
The solution is to compile the $\lambda$-term with its environment by calling
the evaluator at the current stage. The result is generating a next-stage
Scala function. The evaluator is passed as @ev@ to the method @close@.
The following code implements the idea:
\begin{lstlisting}
  type ValSt = (Value, Store)
  def emit_compiled_clo(f: (Rep[Value], Rep[Store]) => Rep[ValSt], λ: Lam, ρ: Exp[Env]): Rep[Value]
  def close(ev: Expr => Ans)(λ: Lam, ρ: Rep[Env]): Rep[Value] = {
    val Lam(x, e) = λ
    val f: (Rep[Value], Rep[Store]) => Rep[ValSt] = { case (v: Rep[Value], σ: Rep[Store]) =>
        val α = alloc(σ, x)
        ev(e)(ρ + (unit(x) → α))(σ + (α → v))
    }; emit_compiled_clo(f, λ, ρ)
  }
\end{lstlisting}

@@

Now, compared with the unstaged interpreter, we observe a key difference of where
we use the evaluation function @ev@.  In the unstaged interpreter, the
invocation of @ev@ happens in @ap_clo@, i.e., at the application time; while in
the staged interpreter, we eagerly apply @ev@ when denoting the $\lambda$-terms to
our next-stage value domains, i.e., at the value-representation time.

### 2019-09-11 paper/40_absinterp.tex
lambdas

It is important that the abstract stores now map addresses to sets of abstract
values, indicating that an address may point to an over-approximated set of
runtime values.  This can be justified by the approximation nature and
nondeterminism during the analysis.  For example, when analyzing a
conditional expression, we may not have enough information to decide which
branch will be taken, thus a sound treatment is to explore both of the branches.
Also, at some later point, we need to compute the join of two path.  Another
nondeterminism comes from function applications, for instance, @f(a)@, the
possible $\lambda$ terms of @f@ are computed approximately.  As the
result, we need to store all the possible callees, as well as we will
retrieve a set of closures when dereferencing an address.

@@

The value representation of lambda terms is still a defunctionalized closure of
type @CloV@, but we lift it to a singleton set of @AbsValue@ (remember that
type @Value@ is an alias of @Set[AbsValue]@).
\begin{lstlisting}
  def close(ev: Expr => Ans)(λ: Lam, ρ: Env): Value = Set(CloV(λ, ρ))
\end{lstlisting}

>>>>

It is important that the abstract stores now map addresses to sets of abstract
values, indicating that an address may point to an over-approximated set of
runtime values.  This can be justified by the approximation nature and
nondeterminism during the analysis.  For example, when analyzing a
conditional expression, we may not have enough information to decide which
branch will be taken, thus a sound treatment is to explore both of the branches.
Also, at some later point, we need to compute the join of two path.  Another
nondeterminism comes from function applications, for instance, @f(a)@, the
possible $\lambda$-terms of @f@ are computed approximately.  As the
result, we need to store all the possible callees, as well as we will
retrieve a set of closures when dereferencing an address.

@@

The value representation of $\lambda$-terms is still a defunctionalized closure of
type @CloV@, but we lift it to a singleton set of @AbsValue@ (remember that
type @Value@ is an alias of @Set[AbsValue]@).
\begin{lstlisting}
  def close(ev: Expr => Ans)(λ: Lam, ρ: Env): Value = Set(CloV(λ, ρ))
\end{lstlisting}

### 2019-09-11 paper/50_stageabs.tex
lambdas

\paragraph{Primitive Operations} 
When deriving the staged concrete interpreter from its unstaged counterpart, we
notice that the recursive calls of the evaluation function @ev@ on the lambda
body are shifted from @ap_clo@ to @close@. In other words, we eagerly
specialize the interpreter and generate code every time when we reach a
$\lambda$ term, instead of lazily calling @ev@ when the applications happen.
Similarly, here we use the same way to handle staged $\lambda$ terms.  Additionally,
in the staged abstract interpreter, we have to handle nondeterminism incurred
by the over-approximation of runtime behavior.  In the rest of this part, we
keep our roadmap and focus on discussing closure representations and function
applications for the staged version.

>>>>

\paragraph{Primitive Operations} 
When deriving the staged concrete interpreter from its unstaged counterpart, we
notice that where the evaluation @ev(e)@ is invoked are shifted from @ap_clo@ to
@close@, where @e@ is the body expression of a $\lambda$-term.
In other words, when staging, we eagerly specialize the interpreter and
generate code every time when we reach a $\lambda$-term, instead of lazily
calling @ev@ when the applications happen.  Similarly, here we use the same way
to handle staged $\lambda$-terms.  Additionally, in the staged abstract
interpreter, we have to handle nondeterminism incurred by the
over-approximation of runtime behavior.  In the rest of this part, we keep our
roadmap and focus on discussing closure representations and function
applications for the staged version.

### 2019-09-11 paper/60_cases.tex
lambdas

As we have already seen in Section \ref{sai}, the primitive operation @close@
can denote a $\lambda$ term to a next-stage Scala function. Similar to
compilers, the specializing of the staged abstract interpreter to a $\lambda$
term is also independent of other parts of the analyzed program.
\begin{lstlisting}
  def close(ev: Eval => Ans)(λ: Expr, ρ: Rep[Env]): Rep[Value]
\end{lstlisting}

>>>>

As we have already seen in Section \ref{sai}, the primitive operation @close@
can denote a $\lambda$-term to a next-stage Scala function. Similar to
compilers, the specializing of the staged abstract interpreter to a
$\lambda$-term is also independent of other parts of the analyzed program.
\begin{lstlisting}
  def close(ev: Eval => Ans)(λ: Expr, ρ: Rep[Env]): Rep[Value]
\end{lstlisting}

### 2019-09-11 paper/70_eval.tex
lambdas

\vspace{-5pt}
\paragraph{Implementation and Evaluation Environment} We implement the abstract
interpreters for control-flow analysis, i.e., a value-flow analysis that
computes which lambda terms can be called at each call-site.
With other suitable abstract domains, the abstract interpreter
can also be extended to other analyses.
A front-end desugars Scheme to a small core language that makes the analyzer
easy to implement. We implement two common monovariant 0-CFA-like analyses: one
is equipped with store-widening (0CFA-SW for short), the other one is not (0CFA
for short). It is traditionally expected that the store-widened analysis be faster than the
alternative.  Our prototype implementation currently generates Scala code. The
generated Scala code will be compiled and executed on the JVM, though it is possible to
generate C/C++ code in the future. In the experiments, we implement several
optimizations mentioned in Section \ref{staged_ds}, specifically, including the
selective caching scheme and IR-level rewriting rules (for pairs, lists, sets,
and maps) that exploit partially-static data. These optimizations are useful to
reduce the size of generated code and improve performance.

@@

Figure~\ref{evaluation_result} shows the evaluation result, comparing the
performance improved by staging on the two monovariant CFAs. The
\textit{unstaged} and \textit{staged} columns show the median time (excluding
pre-compilation) to finish the analysis. The column
$\frac{\text{unstaged}}{\text{staged}}$ shows the improvement. The dash @-@ in
some cells indicates timeout.  As expected, the staged version significantly
outperforms the unstaged version on all benchmarks.
%We observe that simple value domains used in the benchmark usually leads to
%better speed-ups, for example, the @kcfa-worst@ series only uses lambdas and
%booleans.
For 0CFA without store-widening, we observe an average speedup of ~17 times;
for 0CFA with store-widening, the average speedup is ~7 times.
\texttt{kcfa-worst-64} has the highest speedup among all the tests. Although it
has a comparably large size of AST, it has a rather simple structure which only
involves booleans, lambda terms, and applications. We conjecture that the simple
domain and structure allow the generated code to be efficiently optimized by the
JVM. However, on the corresponding 0CFA-SW, we do not observe such a significant
speedup. Due to the complexity of generated code and next-stage running platform
(JVM), we leave the detailed empirical analysis of which static analysis can
benefit more from the specialization as future work.

>>>>

\vspace{-5pt}
\paragraph{Implementation and Evaluation Environment} We implement the abstract
interpreters for control-flow analysis, i.e., a value-flow analysis that
computes which $\lambda$-terms can be called at each call-site.
With other suitable abstract domains, the abstract interpreter
can also be extended to other analyses.
A front-end desugars Scheme to a small core language that makes the analyzer
easy to implement. We implement two common monovariant 0-CFA-like analyses: one
is equipped with store-widening (0CFA-SW for short), the other one is not (0CFA
for short). It is traditionally expected that the store-widened analysis be faster than the
alternative.  Our prototype implementation currently generates Scala code. The
generated Scala code will be compiled and executed on the JVM, though it is possible to
generate C/C++ code in the future. In the experiments, we implement several
optimizations mentioned in Section \ref{staged_ds}, specifically, including the
selective caching scheme and IR-level rewriting rules (for pairs, lists, sets,
and maps) that exploit partially-static data. These optimizations are useful to
reduce the size of generated code and improve performance.

@@

Figure~\ref{evaluation_result} shows the evaluation result, comparing the
performance improved by staging on the two monovariant CFAs. The
\textit{unstaged} and \textit{staged} columns show the median time (excluding
pre-compilation) to finish the analysis. The column
$\frac{\text{unstaged}}{\text{staged}}$ shows the improvement. The dash @-@ in
some cells indicates timeout.  As expected, the staged version significantly
outperforms the unstaged version on all benchmarks.
%We observe that simple value domains used in the benchmark usually leads to
%better speed-ups, for example, the @kcfa-worst@ series only uses lambdas and
%booleans.
For 0CFA without store-widening, we observe an average speedup of ~17 times;
for 0CFA with store-widening, the average speedup is ~7 times.
\texttt{kcfa-worst-64} has the highest speedup among all the tests. Although it
has a comparably large size of AST, it has a rather simple structure which only
involves booleans, $\lambda$-terms, and applications. We conjecture that the simple
domain and structure allow the generated code to be efficiently optimized by the
JVM. However, on the corresponding 0CFA-SW, we do not observe such a significant
speedup. Due to the complexity of generated code and next-stage running platform
(JVM), we leave the detailed empirical analysis of which static analysis can
benefit more from the specialization as future work.
